A Mechanised Library for Generic (Co)Programming

### 2024-02-22 paper/paper.tex
Moved adjoint folds to the apendix to save space
Types and morphisms

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \mintinline{Coq}{App F A -> B} (or \mintinline{Coq}{B -> App F A}). 
We want the following kind of properties to hold: 
\begin{minted}{Coq}
forall (x y : App C A) (f : App C A -> B), AppR x y -> f x = f y
\end{minted}
Moreover, we want this to be true for any \mintinline{Coq}{B}, and equality.
We address this in the following section.

@@

The need to reason about different forms of equality forces us to couple every
type with an associated equivalence relation. Furthermore, we will only allow
to reason about the equality of \emph{proper morphisms}, i.e. morphisms that
respect the necessary equivalence relations.

\begin{minted}{Coq}
Reserved Notation "f =e g" (at level 70, no associativity).
Class equiv A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).

Reserved Notation "x ~> y" (at level 95, right associativity, y at level 200).
Context `{eA : equiv A} `{eB : equiv B}.

 Structure morph :=
   MkMorph
     { app :> A -> B;
       app_eq : forall x y, x =e y -> app x =e app y
     }.
Arguments morph A {eA} B {eB}.
Notation "x ~> y" := (morph x y).
\end{minted}

\dcas{We provide common morphisms for polynomials: sums, products, const,
exponential (Add examples?)}

By default, we associate every type with Coq's propositional equality, but allow
the use of other forms of equality. In most cases, we will use either
propositional equality, extensional equality, or a mixture of both, e.g. for
container extensions, we use an arbitrary equality for shapes, extensional
equality for the contained elements, but Coq's propositional equality for
positions:
\begin{minted}{Coq}
  Inductive AppR `{Cont Sh P} `{equiv X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
Notice the use of \mintinline{Coq}{=e} for both shapes, and the continuations
of container extensions. 
The reason to use Coq's propositional equality for positions is to avoid a
circularity in the definitions: arbitrary equivalences for positions would
force us to define the contents as morphisms from positions to the contained
elements, but that would as well force us to redefine container extensions so
that they are only applicable to types with arbitrary equalities and morphisms.
We will see in the following section why this is problematic for defining
initial algebras of container extensions.
\dcas{I am actually not sure of what I wrote above, and it may be something I
simply overlooked. I will look into the Coq code and try to generalise it if I
have the time}

\subsubsection{Polynomial Types, Equivalences, and Morphisms}
We associate some types with their standard equivalences:
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
Instance prop_eq : equiv Prop.
Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
\end{minted}

\subsubsection{Isomorphisms with Container Extensions}

@@

\bibliography{references.bib}
\end{document}

>>>>

However, the use of a different equality for container extensions now forces us
to deal with the fact that some types have different definitions of equality.
In particular, we want to reason about the equality of functions of types such
as \mintinline{Coq}{App F A -> B} (or \mintinline{Coq}{B -> App F A}). Since 
these types now come with their own equivalence, any function that manipulates
them needs to be \emph{respectful}. I.e.\ given 
\mintinline{Coq}{R : X -> X -> Prop} and \mintinline{Coq}{R' : Y -> Y -> Prop},
we want functions (morphisms) that satisfy the following property:
\begin{minted}{Coq}
forall (x y : X), R x y -> R' (f x) (f y)
\end{minted}

@@

We address the different forms of equality by coupling every type with an
associated equivalence relation, and every morphism with a proof that it is
\emph{respectful} (i.e. \emph{proper} morphisms with respect to the function
\emph{respectfulness} relation). We use \mintinline{Coq}{=e} to denote the
equivalence relation associated with a type. Every Coq type is associated by
default with the standard Coq propositional equality, unless a different
equivalence is specified. We use the type-class mechanism to help Coq's code
extraction mechanism to remove any occurence of custom equivalence relations
in the extracted OCaml code.
\begin{minted}{Coq}
Reserved Notation "f =e g" (at level 70, no associativity).
Class equiv A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{minted}
Given types \mintinline{Coq}{A} and  \mintinline{Coq}{B}, with their respective
equivalence relations \mintinline{Coq}{eA : A -> A -> Prop} and
\mintinline{Coq}{eB : B -> B -> Prop}, the we define the type
\mintinline{Coq}{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \mintinline{Coq}{R} to \mintinline{Coq}{R'}.
\begin{minted}{Coq}
Structure morph A {eA : equiv A} B {eB : equiv B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{minted}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \mintinline{Coq}{Prop} in the code, so objects of type
\mintinline{Coq}{A ~> B} will be extracted to the OCaml equivalent to
\mintinline{Coq}{A -> B}.

\subsubsection{Polynomial Types}

We define a number of equivalences for polynomial types. These equivalences are
be used when building proper morphisms that manipulate their respective types.
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
Instance prop_eq : equiv Prop.
Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
\end{minted}

Most of the definitions that involve functions and polynomial types are
straightworward.  Identity and composition are defined as 
\mintinline{Coq}{fun x => x} and \mintinline{Coq}{fun f g x => f (g x)} 
respectively, and the proofs that they are proper morphisms is straightforward,
and automatically discharged by Coq. 
Products are built using function \mintinline{Coq}{fun f g x => (f x, g x)},
with the projections being the standard Coq  \mintinline{Coq}{fst} and
\mintinline{Coq}{snd} functions. Similarly, sum injections are encoded using
Coq's \mintinline{Coq}{inl} and \mintinline{Coq}{inr} constructors, and
pattern matching on them uses the function 
\mintinline{Coq}{fun f g x => match x with | inl y => f y | inr y => g y end}.
The proofs that these morphisms are proper are straightforward.  Finally, we
also provide functions for currying/uncurrying, and partially applying a proper
morphism. We avoid using definitions, or unfold them, to help Coq's code
extraction mechanism to inline as many of these small definitions as possible.

These definitions are isomorphic to the equivalent container extension.

We illustrate how these definitions
\paragraph*{Isomorphisms with Container Extensions} Our definitions involving
polynomial types are isomorphic to polynomial functors as container extensions.

\subsubsection{Generalised Rewriting for Morphisms}

@@

\bibliography{references.bib}
\appendix

\end{document}
